---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

**1. Code for reading the dataset and processing the data**

Setting up environment
```{r, tidy=TRUE, hide=TRUE, message=FALSE, warning=FALSE}
rm(list=ls())
library("data.table"); library("dplyr"); library("timeDate"); library("ggplot2")
setwd('~/GitHub/05_ReproducableResearch/RepData_PeerAssessment1')
```

Loading data
```{r, tidy=TRUE}
f <- file.path(getwd(),"activity.zip") %>% unzip()
DT_activity<-data.table(read.csv(file.path(getwd(),"activity.csv")))
```
Making `interval` an index from `1:288` to number the 5-minute intervals in a day.
```{r, tidy=TRUE}
DT_activity[,interval:=c(1:288)]
```


**2. Histogram of the total number of steps taken per day**

Manipulating data
```{r, tidy=TRUE}
DT_activity <- group_by(DT_activity,date) #group data
DT_totalSteps <- summarise(DT_activity, sum(steps)) %>% setnames(c("date","totalSteps")) 
summarise(ungroup(DT_activity), sum(steps)) #ungroup data
```

Creating plot
```{r, tidy=TRUE}
with(DT_totalSteps,
        hist(totalSteps, 
                main="Histogram of Steps Taken Per Day",
                xlab="Number of steps taken per day"
        )
)
```

**3. Mean and median of the total number of steps taken per day**
```{r, tidy=TRUE}
summarise(DT_totalSteps, mean(totalSteps,na.rm=T), median(totalSteps,na.rm=T)) %>% setnames(c("mean","median")) %>% data.table()
```
**4. Time series plot of the average number of steps taken**

Calculating the mean steps taken by interval
```{r, tidy=TRUE}
DT_activity <- group_by(DT_activity,interval) #group data
summarise(DT_activity, mean(steps)) %>% setnames(c("interval","averageSteps")) %>% head(3)
```
Note the values for `mean(steps)` are all NA. This tells us there is at least one incomplete case within each interval. To deal with this we specify `na.rm=TRUE` within the call to `mean()`.
```{r, tidy=TRUE}
DT_activityPattern <- summarise(DT_activity, mean(steps,na.rm=TRUE)) %>% 
  setnames(c("interval","averageSteps"))
head(DT_activityPattern, 3)
summarise(ungroup(DT_activity), mean(steps)) #ungroup data
```
Creating time-series plot 
```{r, tidy=TRUE}
with(DT_activityPattern,
        plot(interval, averageSteps,
                main="Average Daily Activity Pattern",
                xlab="Interval",
                ylab="Average steps",
                type="l"
        )
) 

```
**5. The 5-minute interval that, on average, contains the maximum number of steps**

Calculating the max of `averageSteps` and the interval at which it occurs
```{r, tidy=TRUE}
summarize(DT_activityPattern, which.max(averageSteps), max(averageSteps)) %>% 
  setnames(c("interval","max")) %>% data.table()
```
**6. Code to describe and show a strategy for imputing missing data**

Counting the number of rows that have missing data
```{r, tidy=TRUE}
summarize(DT_activity, sum(is.na(steps))) %>% setnames(c("NAs")) %>% data.table()
```
Filling NA values with the mean
```{r, tidy=TRUE, warning=FALSE}
for (col in c("steps")) DT_activity[is.na(get(col)), (col) := mean(DT_activity$steps,na.rm=TRUE)]
```
Verifying the NA values were filled
```{r, tidy=TRUE}
summarize(DT_activity, sum(is.na(steps))) %>% setnames(c("NAs")) %>% data.table()
```

**7. Histogram of the total number of steps taken each day after the missing data has been filled**

Manipulating data
```{r, tidy=TRUE}
DT_activity <- group_by(DT_activity,date) #group data
DT_totalSteps <- summarise(DT_activity, sum(steps)) %>% setnames(c("date","totalSteps")) 
summarise(ungroup(DT_activity), sum(steps)) #ungroup data
```
Creating plot
```{r, tidy=TRUE}
with(DT_totalSteps,
        hist(totalSteps, 
                main="Histogram of Steps Taken Per Day",
                xlab="Number of steps taken per day"
        )
)
```
Calculating and reporting the mean and median of the total number of steps taken per day
```{r, tidy=TRUE}
summarise(DT_totalSteps, mean(totalSteps),median(totalSteps)) %>% 
  setnames(c("mean","median")) %>% data.table()
```
**Mean and median of the total number of steps taken per day**

These values differ from the estimates from the first part of the assignment. The mean and median decreased by `14.45` steps and `109` steps, respectively. 

**8. Panel plot comparing the average number of steps taken per 5-minute interval accross weekdays and weekends**

Calculating mean steps taken by interval for weekdays and for weedends
```{r, tidy=TRUE}
DT_activity <- DT_activity[,wday:=isWeekday(DT_totalSteps$date, wday=1:5)] #create wday variable
DT_wdays <- DT_activity[wday==TRUE] %>% #filter rows where wday==TRUE
  group_by(interval) %>% #group data
  summarise(mean(steps)) %>% #calculate mean steps by interval
  setnames(c("interval","averageSteps")) %>%
  setorder(interval)
DT_wends <- DT_activity[wday==FALSE] %>% #filter rows where wday==FALSE
  group_by(interval) %>% #group data
  summarise(mean(steps)) %>% #calculate mean steps by interval
  setnames(c("interval","averageSteps")) %>%
  setorder(interval)
```
Creating panel plot
```{r}
ggplot(DT_wdays,aes(interval,averageSteps)) + 
  geom_line(aes(color="wdays")) + 
  geom_line(data=DT_wends,aes(color="wends")) + 
  labs(list(x="Interval", y="Average steps")) + 
  scale_x_continuous(breaks=seq(0,300,50)) +
  scale_y_continuous(breaks=seq(0,250,50)) +
  scale_colour_manual("", breaks = c("wdays", "wends"), values = c("blue", "brown")) + 
  ggtitle("Average Daily Activity Pattern: Weekdays & Weekends") + 
  theme(plot.title = element_text(face="bold"),
        axis.title.x = element_text(face="bold"),
        axis.title.y = element_text(face="bold"),
        legend.position = c(.91, .85))
```