---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

**Setup environment**
```{r, hide=TRUE, message=FALSE}
rm(list=ls())
library("data.table"); library("dplyr"); library("timeDate"); library("R.devices")
setwd('~/GitHub/05_ReproducableResearch/RepData_PeerAssessment1')
```
**Load data**
```{r}
f <- file.path(getwd(),"activity.zip") %>% unzip()
DT_activity<-data.table(read.csv(file.path(getwd(),"activity.csv")))
```
Make `interval` an index from `1:288` to number the 5-minute intervals in a day.
```{r}
DT_activity[,interval:=c(1:288)]
```
## What is mean total number of steps taken per day?
Calculate total steps per day
```{r}
DT_activity <- group_by(DT_activity,date) #group data
DT_totalSteps <- summarise(DT_activity, sum(steps)) %>% setnames(c("date","totalSteps")) 
summarise(ungroup(DT_activity), sum(steps)) #ungroup data
```
Generate histogram of the total number of steps taken per day
```{r}
with(DT_totalSteps,
        hist(totalSteps, 
                main="Histogram of Steps Taken Per Day",
                xlab="Number of steps taken per day"
        )
)
```
Calculate and report the mean and median of the total number of steps taken per day
```{r}
summarise(DT_totalSteps, mean(totalSteps,na.rm=T), median(totalSteps,na.rm=T)) %>% 
  setnames(c("mean","median")) %>% data.table()
```
## What is the average daily activity pattern?
Calculate the mean steps taken by interval
```{r}
DT_activity <- group_by(DT_activity,interval) #group data
summarise(DT_activity, mean(steps)) %>% setnames(c("interval","averageSteps")) %>% head(3)
```
Note the values for `mean(steps)` are all NA. This tells us there is at least one incomplete case within each interval. To fix this we specify `na.rm=TRUE` within the call to `mean()`.
```{r}
DT_activityPattern <- summarise(DT_activity, mean(steps,na.rm=TRUE)) %>% 
  setnames(c("interval","averageSteps"))
head(DT_activityPattern, 3)
summarise(ungroup(DT_activity), mean(steps)) #ungroup data
```
Plot a time series with average steps taken (y-axis) for each 5-minute intervals (x-axis). 
```{r}
with(DT_activityPattern,
        plot(interval, averageSteps,
                main="Average Daily Activity Pattern",
                xlab="Interval",
                ylab="Average steps",
                type="l"
        )
) 

```
**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**  
Calculate the max of `averageSteps` and the interval at which it occurs
```{r}
summarize(DT_activityPattern, which.max(averageSteps), max(averageSteps)) %>% 
  setnames(c("interval","max")) %>% data.table()
```
## Imputing missing values
Count the number of rows containing NA values
```{r}
summarize(DT_activity, sum(is.na(steps))) %>% setnames(c("NAs")) %>% data.table()
```
**First approach**
Fill NA values using the mean
```{r, warning=FALSE}
for (col in c("steps")) DT_activity[is.na(get(col)), (col) := mean(DT_activity$steps,na.rm=TRUE)]
summarize(DT_activity, sum(is.na(steps))) %>% setnames(c("NAs")) %>% data.table()
```
Calculate the total number of steps per day using filled data
```{R}
DT_activity <- group_by(DT_activity,date) #group data
DT_totalSteps <- summarise(DT_activity, sum(steps)) %>% setnames(c("date","totalSteps")) 
summarise(ungroup(DT_activity), sum(steps)) #ungroup data
```
Generate histogram of the total number of steps taken per day
```{r}
with(DT_totalSteps,
        hist(totalSteps, 
                main="Histogram of Steps Taken Per Day",
                xlab="Number of steps taken per day"
        )
)
```
Calculate and report the mean and median of the total number of steps taken per day
```{r}
summarise(DT_totalSteps, mean(totalSteps),median(totalSteps)) %>% 
  setnames(c("mean","median")) %>% data.table()
```
**Do these values differ from the estimates from the first part of the assignment?**
Yes. The mean and median decreased by `14.45` steps and `109` steps, respectively. 

## Are there differences in activity patterns between weekdays and weekends?

Calculate the mean steps taken by interval for weekdays and weedends
```{r}
DT_activity <- DT_activity[,wday:=isWeekday(DT_totalSteps$date, wday=1:5)] #create wday variable
DT_wdays <- DT_activity[wday==TRUE] %>% #filter rows where wday==TRUE
  group_by(interval) %>% #group data
  summarise(mean(steps)) %>% #calculate mean steps by interval
  setnames(c("interval","averageSteps")) %>%
  setorder(interval)
DT_wends <- DT_activity[wday==FALSE] %>% #filter rows where wday==FALSE
  group_by(interval) %>% #group data
  summarise(mean(steps)) %>% #calculate mean steps by interval
  setnames(c("interval","averageSteps")) %>%
  setorder(interval)
```
Create a panel plot comparing the average steps taken per 5-minute interval across weekdays and weekends.
```{r}
with(DT_wdays,
     plot(interval, averageSteps,main="Average Daily Activity Pattern - Weekdays",
          xlab="Interval", ylab="Average steps",
          type="l", col="red"))
with(DT_wends,
     plot(interval, averageSteps,main="Average Daily Activity Pattern - Weekends",
          xlab="Interval", ylab="Average steps",
          type="l", col="blue"))
```
